Cochain level May--Steenrod operations

### 2021-08-30 sec/operads.tex
Improvements after Greg's proofreading

Another groupoid of importance to us is $\mathrm{S} \times \mathrm{S}^{op}$ with covariant functors from it to $\mathbf{Ch}_R$ referred to as \textit{$\mathrm{S}$-bimodules}.
Notice that the inclusions $\mathrm{S} \to \mathrm{S} \times \mathrm{S}^{op}$ induced by $r \mapsto (r,1)$ and $r \mapsto (1,r)$ define forgetful functors
\begin{equation*}
\begin{tikzcd}[column sep=small, row sep=tiny]
& \mathbf{Ch}_R^{\mathrm{S} \times \mathrm{S}^{op}} \arrow[dl, "U_1"', out=-120, in=0] \arrow[dr, "U_2", out=-60, in=180] & \\
\mathbf{Ch}_R^{\mathrm{S}^{op}} & & \mathbf{Ch}_R^{\mathrm{S}},
\end{tikzcd}
\end{equation*}
and that for any object $A$ in $\mathbf{Ch}_R$ the canonical \textit{endomorphism bimodule}
\begin{equation*}
\End_A^A(r, s) = \Hom(A^{\otimes r}, A^{\otimes s})
\end{equation*}
forgets via $U_1$ and $U_2$ to $\End_A$ and $\End^A$ respectively.

@@

A \textit{resolution} in $\mathbf{Ch}_R^\Gamma$ is a morphism $\phi$ of $\Gamma$-modules such that $\phi(r)$ is a resolution in the category of chain complexes of $R[\Gamma_r]$-modules for each $r \in \mathbb{N}$.
A $\Gamma$-module $\mathcal R$ is said to be $E_\infty$ if $\mathcal R(0) = R$ and there exists a resolution $\mathcal R \to \underline{R}$ where $\underline{R}$ is the object in $\mathbf{Ch}_R^\Gamma$ defined by $\underline{R}(r) = R$ and $\underline{R}(\gamma) = \mathrm{id}_R$ for every $r \in \mathbb{N}$ and $\gamma \in \Gamma_r$.

@@

Operads an props are respectively $\mathrm{S}$-modules and $\mathrm{S}$-bimodules enriched with further compositional structure.
These structures are best understood by abstracting the compositional structure naturally present in the coendomorphism $\mathrm{S}$-module $\End^A$, naturally an operad, and the endomorphism $\mathrm{S}$-bimodule $\End_A^A$, naturally a prop.

>>>>

Another groupoid of importance to us is $\mathrm{S} \times \mathrm{S}^{op}$ with covariant functors from it to $\mathbf{Ch}_R$ referred to as \textit{$\mathrm{S}$-bimodules}.
Notice that the inclusions $\mathrm{S} \to \mathrm{S} \times \mathrm{S}^{op}$ induced by $r \mapsto (r,1)$ and $r \mapsto (1,r)$ define forgetful functors
\begin{equation*}
\begin{tikzcd}[column sep=small, row sep=tiny]
& \mathbf{Ch}_R^{\mathrm{S} \times \mathrm{S}^{op}} \arrow[dl, "U_1"', out=-120, in=0] \arrow[dr, "U_2", out=-60, in=180] & \\
\mathbf{Ch}_R^{\mathrm{S}^{op}} & & \mathbf{Ch}_R^{\mathrm{S}}.
\end{tikzcd}
\end{equation*}
Explicitly, $U_1(\mathcal P)(r) = \mathcal P(r,1)$ and $U_2(\mathcal P)(r) = \mathcal P(1,r)$ for any $\mathcal P$ in $\mathbf{Ch}_R^{\mathrm{S} \times \mathrm{S}^{op}}$.
Notice that for any object $A$ in $\mathbf{Ch}_R$ the canonical \textit{endomorphism bimodule}
\begin{equation*}
\End_A^A(r, s) = \Hom(A^{\otimes r}, A^{\otimes s})
\end{equation*}
forgets via $U_1$ and $U_2$ to $\End_A$ and $\End^A$ respectively.

@@

A \textit{resolution} in $\mathbf{Ch}_R^\Gamma$ is a morphism $\phi$ of $\Gamma$-modules such that $\phi(r)$ is a resolution in the category of chain complexes of $R[\Gamma_r]$-modules for each $r \in \mathbb{N}$, where $\Gamma_r$ denotes $\Gamma(r,r)$.
A $\Gamma$-module $\mathcal R$ is said to be $E_\infty$ if $\mathcal R(0) = R$ and there exists a resolution $\mathcal R \to \underline{R}$ where $\underline{R}$ is the object in $\mathbf{Ch}_R^\Gamma$ defined by $\underline{R}(r) = R$ and $\underline{R}(\gamma) = \mathrm{id}_R$ for every $r \in \mathbb{N}$ and $\gamma \in \Gamma_r$.

@@

Operads and props are respectively $\mathrm{S}$-modules and $\mathrm{S}$-bimodules enriched with further compositional structure.
These structures are best understood by abstracting the compositional structure naturally present in the endomorphism $\mathrm{S}$-module $\End_A$ (or $\End^A$), naturally an operad, and the endomorphism $\mathrm{S}$-bimodule $\End_A^A$, naturally a prop.

### 2021-08-30 sec/outlook.tex
Improvements after Greg's proofreading

In physics, Gaiotto, Kapustin, Thorngren \cite{gaiotto2016spin,bhardwaj2017fermionic,kapustin2017fermionic} and others have considered cellular decompositions of spacetime and fields represented by simplicial and cubical cochains.
In order to express subtle interactions between these fields, they have used cup-$i$ products to define relevant action functionals for topological field theories.
We expect that new topological field theories of interest can be defined using Steenrod cup-$(p,i)$ products defined in this work.

>>>>

In physics, Gaiotto, Kapustin, Thorngren \cite{gaiotto2016spin,bhardwaj2017fermionic,kapustin2017fermionic} and others have considered cellular decompositions of spacetime together with fields represented by cellular cochains.
In order to express subtle interactions between these fields, they have used cup-$i$ products to define relevant action functionals for topological field theories.
We expect that new topological field theories of interest can be studied using the Steenrod cup-$(p,i)$ products introduced in this work.

### 2021-08-30 sec/preliminaries.tex
Structural refactoring using new template
Update references and split-off acks
Improvements after Greg's proofreading

A \textit{simplicial set} $X$ is a contravariant functor from the simplex category to the category of sets, and a simplicial map is a natural transformation between two simplicial sets.
As is customary we use the notation
\begin{equation*}
X\big( [n] \big) = X_n, \qquad X(\delta_i) = d_i, \qquad X(\sigma_i) = s_i,
\end{equation*}
and refer to elements in the image of any $s_i$ as \textit{degenerate}.

For each $n \in \mathbb{N}$, the simplicial set $\triangle^n$ is defined by
\begin{equation*}
\triangle^n_k = \Hom_{\triangle} \big( [k], [n] \big), \qquad 
d_i(x) = x \circ \delta_i, \qquad 
s_i(x) = x \circ \sigma_i,
\end{equation*}
and any simplicial set can be expressed as a colimit of these
\begin{equation*}
X = \colim_{\triangle^n \to X} \triangle^n.
\end{equation*}
We represent the non-degenerate elements of $\triangle^n_k$ as an increasing sequence $[v_0, \dots, v_k]$ of non-negative integers each less than or equal to $n$.

@@

A \textit{cubical set} $X$ is a contravariant functor from the cube category to the category of, sets and a cubical map is a natural transformation between two cubical sets.
As is customary, we use the notation
\begin{equation*}
X\big( 2^n \big) = X_n \qquad X(\delta^\varepsilon_i) = d^\varepsilon_i \qquad X(\sigma_i) = s_i,
\end{equation*}
and refer to elements in the image of any $s_i$ as \textit{degenerate}.

For each $n \in \mathbb{N}$, the cubical set $\square^n$ is defined by
\begin{equation*}
\square^n_k = \Hom_{\square} \big( 2^k, 2^n \big), \qquad 
d^\varepsilon_i(x) = x \circ \delta^\varepsilon_i, \qquad 
s_i(x) = x \circ \sigma_i.
\end{equation*}
We represent the non-degenerate elements of $\square^n$ as sequences $x_1 \cdots\, x_n$ with each $x_i \in \big\{[0], [1], [0,1]\big\}$.
For example, $[0][01][1]$ represents $\delta^1 \times \mathrm{id} \times \delta^0$.
Any cubical set can be expressed as a colimit of these
\begin{equation*}
X = \colim_{\square^n \to X} \square^n.
\end{equation*}

>>>>

A \textit{simplicial set} $X$ is a contravariant functor from the simplex category to the category of sets, and a simplicial map is a natural transformation between two simplicial sets.
As is customary, we use the notation
\begin{equation*}
X\big( [n] \big) = X_n, \qquad X(\delta_i) = d_i, \qquad X(\sigma_i) = s_i,
\end{equation*}
and refer to elements in the image of any $s_i$ as \textit{degenerate}.

For each $n \in \mathbb{N}$, the simplicial set $\triangle^n$ is defined by
\begin{equation*}
\triangle^n_k = \Hom_{\triangle} \big( [k], [n] \big), \qquad
d_i(x) = x \circ \delta_i, \qquad
s_i(x) = x \circ \sigma_i,
\end{equation*}
and any simplicial set can be expressed as a colimit of these
\begin{equation*}
X \cong \colim_{\triangle^n \to X} \triangle^n.
\end{equation*}
We represent the non-degenerate elements of $\triangle^n_k$ as increasing sequences $[v_0, \dots, v_k]$ of non-negative integers each less than or equal to $n$.

@@

A \textit{cubical set} $X$ is a contravariant functor from the cube category to the category of sets, and a cubical map is a natural transformation between two cubical sets.
As is customary, we use the notation
\begin{equation*}
X\big( 2^n \big) = X_n \qquad X(\delta^\varepsilon_i) = d^\varepsilon_i \qquad X(\sigma_i) = s_i,
\end{equation*}
and refer to elements in the image of any $s_i$ as \textit{degenerate}.

For each $n \in \mathbb{N}$, the cubical set $\square^n$ is defined by
\begin{equation*}
\square^n_k = \Hom_{\square} \big( 2^k, 2^n \big), \qquad
d^\varepsilon_i(x) = x \circ \delta^\varepsilon_i, \qquad
s_i(x) = x \circ \sigma_i.
\end{equation*}
We represent the non-degenerate elements of $\square^n$ as sequences $x_1 \cdots\, x_n$ with each $x_i \in \big\{[0], [1], [0,1]\big\}$.
For example, $[0][01][1]$ represents $\delta^1 \times \mathrm{id} \times \delta^0$.
Any cubical set can be expressed as a colimit of these
\begin{equation*}
X \cong \colim_{\square^n \to X} \square^n.
\end{equation*}

### 2021-08-30 sec/steenrod.tex
Improvements after Greg's proofreading

\begin{remark} \label{rmk: Deligne conjecture}
	In \cite{GerstenhaberVoronov} an operad morphism $\mathcal{A}ssoc \to \mathcal O$ is referred to as a \textit{multiplication} on $\mathcal O$.
	In this language, a choice of factorization $\phi \circ \iota$ of a May--Steenrod structure on $\mathcal O$ endows it with a \textit{commutative multiplication} $\phi$.
\end{remark}

@@

Let $A$ be equipped with a May--Steenrod structure
\begin{equation*}
\begin{tikzcd}[column sep = normal, row sep = small]
& \mathcal R \arrow[dr, "\phi", out=0] & \\
\mathcal W \arrow[ur, "\iota", out=55, in=180] \arrow[rr, "\psi"] & & \End_A.
\end{tikzcd}
\end{equation*}
We can relate this structure on $A$ to those considered by May in \cite{may1970general} as follows.
The morphism $\phi$ provides $A$ with the structure of a homotopy associative algebra defined by the image in $\Hom(A^{\otimes 2}, A)$ of a representative in $\mathcal R(2)$ of a generator of its $0\th$-homology.
Restricting $\psi$ to arity $r$ defines a map $\theta \colon \mathcal W(r) \otimes A^{\otimes r} \to A$ and the pair $(A, \theta)$ into an object in May's category $\mathfrak{C}(\mathrm C_r, \infty, \Z)$ as presented in Definitions 2.1 op. cit.
Explicitly, this pair is such that $\psi(e_0) \in \End_A(r)$ is $\mathrm C_r$-homotopic to the iterated product $A^{\otimes r} \to A$.
A claim that follows from the iterated product being a representative of a generator of the $0\th$-homology of $\mathcal R(r)$, and $\iota$ being a quasi-isomorphism of $\mathrm C$-modules.
Furthermore, for $r$ equal to a prime $p$, tensoring the integers with $\Fp$ makes the pair $(A, \theta)$ into an object in May's category $\mathfrak{C}(\mathrm C_p, \infty, \mathbb{F}_p)$.
For any object $(A, \theta)$ in this category, Definition 2.2 in \cite{may1970general} defines operations on the mod $p$ homology of $A$, a construction we review below.
In particular, if $A$ is given by the cochains of a space these products agree with Steenrod's original definitions, and for $A$ being the chains on an infinite loop spaces, with those defined by Araki--Kudo and Dyer--Lashof.

@@

Notice that the Steenrod squares are determined by the Steenrod cup-$(2,i)$ products with $\mathbb{F}_2$-coefficients.
These binary operations are known as cup-$i$ products \cite{steenrod1947products, medina2021newformulas}.
In a similar way, the operations $P$ and $\beta P$ defined below for odd primes are determined by the Steenrod cup-$\big(p, k(p-1)-\varepsilon\big)$ products for $\varepsilon \in \{0,1\}$.
We can explain the appearance of these specific Steenrod cup-$(p,1)$products as follows.
The increase on the degree of a $q$-cycle after applying $D^p_{k(p-1)-\varepsilon}$ to it is $(p-1)(q+k) - \varepsilon$, which can be rewritten as $2t(p-1) - \varepsilon$ if $q$ is even an $(2t+1)(p-1) - \varepsilon$ if $q$ is odd.
According to Lemma \ref{lem: Thom's theorem}, these are the only homologically non-trivial cases.

@@

\begin{proof}
	As described after Definition~\ref{def: Steenrod products}, for any prime $p$ the pair $(A, \theta)$ is an object in May's category $\mathfrak{C}(\mathrm C_p, \infty, \mathbb{F}_p)$.
	Furthermore, since we are demanding a factorization $\phi \circ \iota$ with $\phi$ an operad map from an $E_\infty$-operad to $\End_A$, this is an \textit{Adem object} in the sense of Definition 4.1 of \cite{may1970general}.
	The statement presented here is stated and proven as part of Theorem 4.7 loc. cit.
\end{proof}

So far we have considered $\mathrm{C}$-modules, operads and related structures over the category of chain complexes.
It is also useful to consider them over the category of coalgebras, that is to say, requiring each chain complex to be equipped with a coproduct and all structure maps to be morphisms of coalgebras.
As described in Definition~1.2 of \cite{may1970general}, the $\mathrm{C}$-module $\mathcal W$ lifts to this category.
A \textit{comultiplicative May--Steenrod structure} on an operad $\mathcal O$ is a morphism of $\mathrm{C}$-modules $\psi \colon \mathcal W \to \mathcal O$ for which there exists a factorization through an $E_\infty$-operad over the category of coalgebras
\begin{equation*}
\begin{tikzcd}[column sep = normal, row sep = small]
& \mathcal R \arrow[dr, "\phi", out=0] & \\
\mathcal W \arrow[ur, "\iota", out=55, in=180] \arrow[rr, "\psi"] & & \mathcal O
\end{tikzcd}
\end{equation*}
such that $\iota$ is a quasi-isomorphism over the category of coalgebras and $\phi$ is a morphism of operads.

@@

\begin{proof}
	As described after Definition~\ref{def: Steenrod products}, for any prime $p$ the pair $(A, \theta)$ is an object in May's category $\mathfrak{C}(\mathrm C_p, \infty, \mathbb{F}_p)$.
	Furthermore, since we are demanding a factorization $\phi \circ \iota$ with $\iota$ a quasi-isomorphism in the category of $\mathrm{C}$-modules over the category of coalgebras, this is a \textit{Cartan object} as defined in pag. 161 of \cite{may1970general}.
	The statement presented here is stated in pag.~165 loc.~cit.
\end{proof}

For the even prime, constructive proofs at the cochain level of the Adem and Cartan relations have been given respectively in \cite{medina2021adem} and \cite{medina2020cartan}.

>>>>

\begin{remark} \label{rmk: Deligne conjecture}
	In \cite{GerstenhaberVoronov} an operad morphism $\mathcal{A}ssoc \to \mathcal O$ is referred to as a \textit{multiplication} on $\mathcal O$.
	In this language, a choice of factorization $\phi \circ \iota$ of a May--Steenrod structure on $\mathcal O$ endows it with $\phi$ a \textit{multiplication commutative up to coherent homotopies}.
\end{remark}

@@

Let $A$ be equipped with a May--Steenrod structure
\begin{equation*}
\begin{tikzcd}[column sep = normal, row sep = small]
& \mathcal R \arrow[dr, "\phi", out=0] & \\
\mathcal W \arrow[ur, "\iota", out=55, in=180] \arrow[rr, "\psi"] & & \End_A.
\end{tikzcd}
\end{equation*}
We can relate this structure on $A$ to those considered by May in \cite{may1970general} as follows.
The morphism $\phi$ provides $A$ with the structure of a homotopy associative algebra defined by the image in $\Hom(A^{\otimes 2}, A)$ of a representative in $\mathcal R(2)$ of a generator of its $0\th$-homology.
Restricting $\psi$ to arity $r$ defines a map $\theta \colon \mathcal W(r) \otimes A^{\otimes r} \to A$ that makes the pair $(A, \theta)$ into an object in May's category $\mathfrak{C}(\mathrm C_r, \infty, \Z)$ as presented in Definitions 2.1 op. cit.
Explicitly, this means that the pair is such that $\psi(e_0) \in \End_A(r)$ is $\mathrm C_r$-homotopic to the iterated product $A^{\otimes r} \to A$.
A claim that follows from the iterated product being a representative of a generator of the $0\th$-homology of $\mathcal R(r)$, and $\iota$ being a quasi-isomorphism of $\mathrm C$-modules.
Furthermore, for $r$ equal to a prime $p$, tensoring the integers with $\Fp$ makes the pair $(A, \theta)$ into an object in May's category $\mathfrak{C}(\mathrm C_p, \infty, \mathbb{F}_p)$.
For any object $(A, \theta)$ in this category, Definition~2.2 in \cite{may1970general} defines operations on the mod $p$ homology of $A$, a construction we review below.
In particular, if $A$ is given by the cochains of a space these products agree with Steenrod's original definitions, and for $A$ being the chains on an infinite loop spaces, with those defined by Araki--Kudo and Dyer--Lashof.

@@

Notice that the Steenrod operations above, corresponding to Steenrod squares in the context of spaces, are determined by the Steenrod cup-$(2,i)$ products with $\mathbb{F}_2$-coefficients.
These binary operations are known as cup-$i$ products \cite{steenrod1947products, medina2021newformulas} in the space context.
In a similar way, the operations $P$ and $\beta P$ defined below for odd primes are determined by the Steenrod cup-$\big(p, k(p-1)-\varepsilon\big)$ products for $\varepsilon \in \{0,1\}$.
We can explain the appearance of these specific Steenrod cup-$(p,1)$ products as follows.
The increase on the degree of a $q$-cycle after applying $D^p_{k(p-1)-\varepsilon}$ to it is $(p-1)(q+k) - \varepsilon$, which can be rewritten as $2t(p-1) - \varepsilon$ if $q$ is even and $(2t+1)(p-1) - \varepsilon$ if $q$ is odd.
According to Lemma \ref{lem: Thom's theorem}, these are the only homologically non-trivial cases.

@@

\begin{proof}
	As described after Definition~\ref{def: Steenrod products}, for any prime $p$ the pair $(A, \theta)$ is an object in May's category $\mathfrak{C}(\mathrm C_p, \infty, \mathbb{F}_p)$.
	Furthermore, since we are demanding a factorization $\phi \circ \iota$ with $\phi$ an operad map from an $E_\infty$-operad to $\End_A$, the pair $(A, \theta)$ is an \textit{Adem object} in the sense of Definition 4.1 of \cite{may1970general}.
	The statement presented here is stated and proven as part of Theorem 4.7 loc. cit.
\end{proof}

So far we have considered $\mathrm{C}$-modules, operads and related structures over the category of chain complexes.
It is also useful to consider them over the category of coalgebras, that is to say requiring each chain complex to be equipped with a coproduct and all structure maps to be morphisms of coalgebras.
As described in Definition~1.2 of \cite{may1970general}, the $\mathrm{C}$-module $\mathcal W$ lifts to this category.
A \textit{comultiplicative May--Steenrod structure} on an operad $\mathcal O$ is a morphism of $\mathrm{C}$-modules $\psi \colon \mathcal W \to \mathcal O$ for which there exists a factorization through an $E_\infty$-operad over the category of coalgebras
\begin{equation*}
\begin{tikzcd}[column sep = normal, row sep = small]
& \mathcal R \arrow[dr, "\phi", out=0] & \\
\mathcal W \arrow[ur, "\iota", out=55, in=180] \arrow[rr, "\psi"] & & \mathcal O
\end{tikzcd}
\end{equation*}
such that $\iota$ is a quasi-isomorphism over the category of coalgebras and $\phi$ is a morphism of operads.

@@

\begin{proof}
	As described after Definition~\ref{def: Steenrod products}, for any prime $p$ the pair $(A, \theta)$ is an object in May's category $\mathfrak{C}(\mathrm C_p, \infty, \mathbb{F}_p)$.
	Furthermore, since we are demanding a factorization $\phi \circ \iota$ with $\iota$ a quasi-isomorphism in the category of $\mathrm{C}$-modules over the category of coalgebras, this is a \textit{Cartan object} as defined in \cite[p.161]{may1970general}.
	The statement presented here is stated in page~165 loc.~cit.
\end{proof}

For the even prime, effective proofs at the cochain level of the Adem and Cartan relations have been given respectively in \cite{medina2021adem} and \cite{medina2020cartan}.
Explicitly, these proof construct cochains whose coboundaries descends to the relations in cohomology.
